---
title: "chap10-ipw"
author: "yousri elfattah"
date: "2024-10-02"
output: word_document
---

# Chapter 10

# Causal Inference with Observational Data

## Inverse Probability Weighting (IPW) and Propensity Score Matching (PSM)

1.  **Fit the Propensity Score Model**: Use logistic regression to calculate propensity scores (the probability of treatment given covariates).
2.  **Calculate Inverse Probability Weights**: Inverse the propensity scores to obtain weights.
3.  **Fit the Weighted Outcome Model**: Use these weights in the outcome regression model to estimate the ATE.

### Loading the dataset:

```{r}
# Load necessary library
library(haven)  # for read_dta function

# Load the dataset
data <- haven::read_dta("cattaneo2.dta")

# Convert to standard data frame for full column display
data_df <- as.data.frame(data)

# Inspect the first few rows and column names
head(data_df)

# Check the structure of the dataset to see data types
# Use capture.output to capture the output of str(data)
str_output <- capture.output(str(data))
writeLines(str_output, "str_data_output.txt")

# Print to console to verify
cat("Output of str(data) saved to str_data_output.txt\n")
```

### Initialize the CausalModel

```{r}
library(CausalModels)
library(survey)

# Convert relevant columns to factors
data$mbsmoke <- as.factor(data$mbsmoke)
data$mage <- as.factor(data$mage)
confounders <- c("mmarried", "prenatal1", "mage", "fbaby")

# Initialize the parameters for CausalModels package
CausalModels::init_params(
  outcome = "bweight", 
  treatment = "mbsmoke", 
  covariates = confounders, 
  data = data
)


```

### Estimating Average Treatment Effect:

```{r}
# Fit the propensity score model
propensity_model <- glm(
  formula = mbsmoke ~ mage + mmarried + I(mmarried * mmarried) + prenatal1 + I(prenatal1 * prenatal1) + fbaby + I(fbaby * fbaby),
  family = binomial(link = "logit"), 
  data = data
)

# Calculate propensity scores
propensity_scores <- predict(propensity_model, type = "response")

# Compute inverse probability weights
data$weights <- ifelse(data$mbsmoke == 1, 1 / propensity_scores, 1 / (1 - propensity_scores))

# Fit the weighted outcome model using survey package
outcome_model <- svyglm(
  formula = bweight ~ mbsmoke + mage + mmarried + (mbsmoke * mmarried) + I(mmarried * mmarried) + prenatal1 + 
            (mbsmoke * prenatal1) + I(prenatal1 * prenatal1) + fbaby + (mbsmoke * fbaby) + I(fbaby * fbaby),
  design = svydesign(ids = ~1, weights = ~weights, data = data)
)

# Summary of the weighted outcome model
summary(outcome_model)

# Extract ATE
ate_ipw <- coef(outcome_model)["mbsmoke1"]
se_ipw <- sqrt(diag(vcov(outcome_model)))["mbsmoke1"]

# Calculate Confidence Intervals
ci_low <- ate_ipw - 1.96 * se_ipw
ci_high <- ate_ipw + 1.96 * se_ipw

# Print Average Treatment Effect (ATE)
cat("Estimate - ", ate_ipw, "\n")
cat("SE       - ", se_ipw, "\n")
cat("95% CI   - (", ci_low, ", ", ci_high, ")\n")
```

## Balance Plots

Creating balance plots to compare the distribution of covariates between treated and control groups, both before and after matching or weighting, is essential to check the quality of propensity score matching or weighting. This can be done using the MatchIt package in R for propensity score matching and the cobalt package for balance plots.

### Steps to Create Balance Plots:

1.  **Fit the Propensity Score Model**: Compute the propensity scores.
2.  **Apply Matching/Weighting**: Use MatchIt to perform matching.
3.  **Balance Plots**: Use cobalt to create balance box plots for covariates.

```{r}
# Load necessary libraries
library(MatchIt)
library(ggplot2)
library(tidyverse)

# Assuming 'data' is your dataset
data$mbsmoke <- as.factor(data$mbsmoke)

# Perform propensity score matching
psm <- matchit(mbsmoke ~ mmarried + prenatal1 + mage + fbaby,
               data = data,
               method = "nearest",
               distance = "logit")

# Extract the matched data, specifying a different name for weights
matched_data <- match.data(psm, weights = "matching_weights")

# Extract common columns for raw and matched data
common_columns <- intersect(names(data), names(matched_data))

# Subset data to have the same columns
data_common <- data[common_columns]
matched_data_common <- matched_data[common_columns]

# Add Group column for comparison
data_common$Group <- "Raw"
matched_data_common$Group <- "Matched"

# Combine both datasets
combined_data <- rbind(data_common, matched_data_common)

# Function to create normalized PMF plots for covariates
create_pmf_plots <- function(combined_data, covariate) {
  # Create a subset of data for plotting
  plot_data <- combined_data %>%
    select(Group, mbsmoke, all_of(covariate)) %>%
    gather(key = "Variable", value = "Value", -Group, -mbsmoke)
  
  # Convert Factors to Numeric for PMF
  if (is.factor(plot_data$Value)) {
    plot_data$Value <- as.numeric(as.factor(plot_data$Value))
  }
  
  # Calculate normalized counts within each group and treatment status
  plot_data <- plot_data %>%
    group_by(Group, mbsmoke, Value) %>%
    summarize(Count = n()) %>%
    group_by(Group, mbsmoke) %>%
    mutate(Probability = Count / sum(Count)) %>%
    ungroup()
  
  # Create PMF plots
  plot <- ggplot(plot_data, aes(x = factor(Value), y = Probability, fill = mbsmoke)) +
    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
    facet_grid(Group ~ mbsmoke, scales = "free_x") +
    theme_minimal() +
    labs(title = paste("Normalized PMF Plot for", covariate), x = covariate, y = "Probability")
  
  print(plot)
}

# List of covariates
covariates <- c("mmarried", "prenatal1", "mage", "fbaby")

# Loop to create and print PMF plots for all covariates
for (covariate in covariates) {
  create_pmf_plots(combined_data, covariate)
}
```

### Session Info

```{r}
sessionInfo()
```
