---
title: "chap10-standardization"
author: "yousri elfattah"
date: "2024-10-08"
output: word_document
---

# Chapter 10

# Causal Inference with Observational Data

## Standardization Model

**Load required libraries**.

The CausalModels library is used to fit causal models like standardization.

```{r}
# Load libraries
library(dplyr)
library(haven)
library(CausalModels)
library(ggplot2)
```

## Loading the dataset:

This code will output the first few rows and the structure of your dataset into text files, making it easier for you to read and review the results.

1.  **Loading the Dataset**:
    -   `read_dta("cattaneo2.dta")` reads the Stata dataset into R.
2.  Viewing the first few rows of data:
    -   as.data.frame(data) converting data to a standard data frame for full column display.
    -   head(data_df) Inspect the first few rows and column names.
3.  **Capturing str(data) Output**:
    -   `capture.output(str(data))` captures the output of the str(data) function as a character vector.
    -   `writeLines(str_output, "str_data_output.txt")` writes this output to a text file named str_data_output.txt.
4.  **Printing Confirmation**:
    -   `cat` statements print confirmation messages to the console, letting you know the outputs have been successfully saved.

```{r}
# Load necessary library
library(haven)  # for read_dta function

# Load the dataset
data <- haven::read_dta("cattaneo2.dta")

# Convert to standard data frame for full column display
data_df <- as.data.frame(data)

# Inspect the first few rows and column names
head(data_df)

# Check the structure of the dataset to see data types
# Use capture.output to capture the output of str(data)
str_output <- capture.output(str(data))
writeLines(str_output, "str_data_output.txt")

# Print to console to verify
cat("Output of str(data) saved to str_data_output.txt\n")
```

## Initialize CausalModels Package

The function `CausalModels::init_params` must be run first before any other function can run.

Arguments:

1.  **Outcome** - the outcome variable (must be continuous). We set outcome = bweight, which is the "infant birthweight (grams)"
2.  **Treatment**—the treatment with the causal effect of interest on the outcome. We set treatment = mbsmoke, which is "1 if the mother smoked." It has a value of one if the mother is a smoker and 0 if she is a nonsmoker.
3.  **Covariates** - a list/vector of covariate names for confounding adjustment. We set the confounders to the list of variables as follows:
    -   **"mmarried"**: "[1 if mother married]{.underline}" takes the value 0 indicating "notmarried", and one indicating "married." 
    -   **"prenatal1"**: "[1 if first prenatal visit in 1 trimester]{.underline}" takes the value "Yes"-"No"
    -   **"fbaby"**: "[1 if first baby]{.underline}" takes the value "Yes"-"No"
    -   **"medu"**: "[mother's education attainment]{.underline}"
4.  **Data** - is the data frame containing the variables in the model.
5.  **Simple** - a boolean indicator to build a default formula with interactions. If true, interactions will be excluded. If false, interactions will be included. By default, simple is set to false.

The line

`data$mbsmoke <- as.factor(data$mbsmoke)`

converts the `data$mbsmoke` variable in the **cattaneo2** dataset into a factor. The mbsmoke variable indicates whether a mother smoked or not. By converting it to a factor, you inform R that this variable should be treated as categorical, which is crucial for models that handle categorical predictors differently than continuous variables. It allows R to properly interpret **mbsmoke** as levels of a categorical variable in the regression model you are running. The `as.factor` function converts a variable into a factor, an R data type used to represent categorical data.

```{r}
# Initialize the parameters for the CausalModels package
data$mbsmoke <- as.factor(data$mbsmoke)
confounders <- c("mmarried", "prenatal1", "fbaby","medu")
CausalModels::init_params(outcome = bweight, treatment = mbsmoke, covariates = confounders, data)
```

## Fit standardization model

Purpose: Fit a standardization model to the dataset data. data: The dataset containing your variables of interest. n.boot = 100: Specifies 100 bootstrap samples to calculate standard errors. simple = TRUE: Indicates that a simplified version of the model should be fitted.

```{r}
# Fit standardization model
std_model <- standardization(data, n.boot = 100, simple = T)
std_model

```

## Using the coefplot package to visualize the coefficients along with their 95% confidence intervals:

**Explanation**:

1.  **Load Packages**: Ensure both **coefplot** and **broom** are installed and loaded.

2.  **Creating Tidy Data Frame**: Using `broom::tidy()` on your model object keeps it clean, but now, it isn't explicitly required by coefplot as coefplot can directly use the model object. For an introduction to broom, see <https://cran.r-project.org/web/packages/broom/vignettes/broom.html>

3.  **Plotting with coefplot**:

    -   Directly pass the model object to **coefplot**.

    -   `intercept=TRUE` to include the intercept term in the plot.

    -   `outerCI=1.96` to define 95% confidence intervals.

```{r}
# Install and load necessary packages
#install.packages("coefplot")
#install.packages("broom")
library(coefplot)
library(broom)

# Create a tidy data frame for the model coefficients
coef_df <- tidy(std_model$model)

# Use coefplot for visualization
coefplot(std_model$model, intercept=TRUE, outerCI=1.96) +
  ggtitle("Standardization Model Coefficients with 95% Confidence Intervals") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  )
```

## Print Summary of Average Treatment Effect (ATE)

Print the summary of the Average Treatment Effect (ATE) estimated by the model.

```{r}
print(std_model$ATE.summary)

```

### Session Info

```{r}
sessionInfo()
```
