---
title: "R Notebook"
output: html_notebook
---

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

The code chunk appears:
```{r}
if(!require('relations')) {
  install.packages('relations')
  library('relations')
}
```






## Relation r1(x, y, z)
```{r}
r1 <-
  data.frame(
    x = c(1, 1, 2, 1),
    y = c(2, 3, 1, 4),
    z = c(1, 2, 4, 2)
  )
r1 <- as.relation(r1)
relation_table(r1)
```
## the selection operation
```{r}
relation_table(relation_selection(relation_selection(r1, x==1), y<4))
```
## the projection operation
```{r}
relation_table(relation_projection(r1, c("x", "y")))
```
## Relation r2(x,z,w)
```{r}
r2 <-
  data.frame(
    x = c(1, 2, 1, 1),
    z = c(2, 4, 2, 4),
    w = c(3, 1, 2, 2)
  )
r2 <- as.relation(r2)
relation_table(r2)
```
```{r}
relation_table(r1 %|><|% r2)



```
```{r}
xy <-
  data.frame(
    x = c(0, 1, 2),
    y = c(1, 3, 5)
  )
xy <- as.relation(xy)
relation_table(xy)

```
```{r}
yz <-
  data.frame(
    y = c(1, 3, 5),
    z = c(2, 8, 4)
  )
yz <- as.relation(yz)
relation_table(yz)
```
```{r}
zx <-
  data.frame(
    z = c(2, 8, 4),
    x = c(1, 2, 0)
  )
zx <- as.relation(zx)
relation_table(zx)
```
```{r}
relation_table(xy %|><|% yz )
relation_table(yz %|><|% zx )
relation_table(xy %|><|% yz %|><|% zx)
```
```{r}
require("sets")
R <- relation(domain = c("a0", "b0", "b1", "c0", "c1", "d0", "d1", "e0", "e1"),
              graph = set(c("d0", "b1"), c("d1", "b0"), c("d1", "b1"),
                          c("d1", "e1"), c("d1", "e0"), c("d0", "e1"),
                          c("a0", "b0"), c("a0", "b1"),
                          c("b0", "c1"), c("b1", "c1"), c("b1", "c0")))
plot(R)
relation_properties(R)
```
```{r}
require("sets")
rel <- relation(domain = c("A", "B", "C", "D", "E", "F"),
              graph = set(c("A", "B"), c("A", "C"), c("B", "C"), c("C", "F"),
                          c("B", "D"), c("B", "E"), c("D", "E"),
                          c("E", "F"), c("D", "F")))
plot(rel)
relation_is_acyclic(rel)
```

Now, click the **Run** button on the chunk toolbar to [execute](https://www.jetbrains.com/help/pycharm/r-markdown.html#run-r-code) the chunk code. The result should be placed under the chunk.
Click the **Knit and Open Document** to build and preview an output.
