---
title: "R Notebook"
output: html_notebook
---

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

The code chunk appears:
```{r}
library(HyperG)

hg1 <- hypergraph_from_edgelist(list(c("A","B","C"),c("A","C","E"),c("A","B","F","G"),c("A","B","D")))
is.hypertree(hg1)
```

Type any R code in the chunk, for example:
```{r}
plot(hg1)
```
```{r}
summary(hg1)
```
```{r}
library(gRbase)
g <- as.graph(hg1)
jt <- jTree(g)
jt
```
```{r}
plot(jt)
```

Now, click the **Run** button on the chunk toolbar to [execute](https://www.jetbrains.com/help/pycharm/r-markdown.html#run-r-code) the chunk code. The result should be placed under the chunk.
Click the **Knit and Open Document** to build and preview an output.
