---
title: "R Notebook"
output: html_notebook
---

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

The code chunk appears:
```{r}
library(HyperG)

hg1 <- hypergraph_from_edgelist(list(c("A","E","F"),c("A","B","C"),c("A","C","E"),c("C","D","E")))
plot(hg1)
```
```{r}
hg1 <- hypergraph_from_edgelist(list(c("A","E","F"),c("A","B","C"),c("A","C","E"),c("C","D","E")))
summary(hg1)

```
```{r}
library(gRbase)
hg1 <- hypergraph_from_edgelist(list(c("A","E","F"),c("A","B","C"),c("A","C","E"),c("C","D","E")))
g <- as.graph(hg1)
jt <- jTree(g)
jt
plot(jt)
```
```{r}
hg1 <- hypergraph_from_edgelist(list(c("A","E","F"),c("A","B","C"),c("A","C","E"),c("C","D","E")))
# is.hypergraph(hg1)
# is.chordal(hg1)
is.hypertree(hg1)
```
```{r}

```
```{r}

```


Type any R code in the chunk, for example:
```{r}

is.hypergraph(hg1)
# hg2 <- induced_hypergraph(hg1)
# is.hypertree(hg1)
is.tree(hg1)
```
```{r}
g <- as.graph(hg1)
jt <- jTree(g)
jt
plot(jt)
```
```{r}



```
```{r}
hg2 <- hypergraph_from_edgelist(list(c("A","B","C"),c("A","C","E"),c("A","B","F","G"),c("AC","B","D")))
# is.hypergraph(hg1)
# is.chordal(hg1)
is.hypertree(hg2)
```
```{r}

```
```{r}

```

Now, click the **Run** button on the chunk toolbar to [execute](https://www.jetbrains.com/help/pycharm/r-markdown.html#run-r-code) the chunk code. The result should be placed under the chunk.
Click the **Knit and Open Document** to build and preview an output.
